A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-11-04 paper/main.tex
address minor comments on draft

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Today more than 200 strategies exist in the literature and several
tournaments, excluding Axelrod's, have been undertaken~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012}.

@@

Another set of IPD strategies that have received a lot of attention, the
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire'', are the zero-determinants~\cite{Press2012}. By
forcing a linear relationship between the payoffs the zero-determinant
strategies can ensure that they will never receive less than their opponents.
Zero-determinant strategies are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including zero-determinants, was ran. None of the zero-determinant
strategies ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only a few selected pieces of work have been discussed, the IPD is
rich, and new strategies and competitions are being published every year.
The question, however, still remains the same: what is the best way to play the
game?

This manuscript uses an open source package called
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments, using as many strategies as possible from the literature. The aim
is to evaluate the performance of 194 strategies in IPD tournaments, and
futhermore, explore the factors of their success. This is done not only for
standard round robin tournaments, but also for tournaments with noisy and
tournaments with a probabilistic ending. The different tournament types as well
as the data collection are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performed strategies
for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}.

This manuscripts uses several parameters. These are introduced in the following
sections, however, the full set of parameters and their definitions are given in
Appendix~\ref{app:parameters}. %TODO: fix table

@@

The result summary, Table~\ref{table:output_result}, has \(N\) number of rows
because each row contains information for each strategy that participated in the
tournament. The information include the strategy's rank, median score, the rate
with which the strategy cooperated $(C_r)$, it's wins and the probability that
the strategy cooperated in the opening move. Moreover, the rates of a strategy
being in any of the four states ($CC, CD, DC, DD$), and the rate of which the
strategy cooperated after each state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h!]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result of a signle tournament.}\label{table:output_result}
\end{table}

@@

Each strategy participated in multiple tournaments of the same type (on average 5180).
For example Tit For Tat has participated in a total of 5114
tournaments of each type. The strategy's normalised rank distribution in these
is given in Figure~\ref{fig:tit_for_tat_r_distribution}. Because of the strategies'
multiple entries their performance is evaluated based on the
\textbf{median normalised rank} denoted as \(\bar{r}\). A value of \(\bar{r} =
0\) corresponds to a strategy winning the tournament where a a value of
\(\bar{r} = 1\) corresponds to the strategy coming last.

@@

The top strategies in noisy tournaments include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy,
$e$, $\pi$ and $\phi$ (all from~\cite{axelrodproject}). The Retaliate and
Limited Retaliate strategies are implemented in~\cite{axelrodproject} by the
same contributor. The strategies are designed to defect if the opponent has
tricked the strategy more often than \(x\%\) of the times that the strategy has
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are skewed both towards 0
and 1. This indicates that although the top ranked strategies mainly performed
well, there are several tournaments that they ranked in the bottom half.

@@

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinformed learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever are ranked
$14^{\text{th}}$ and $15^{\text{th}}$. These strategies are actually the same;
they will defect until their opponent defect, then they will cooperate until the
end. Both strategies have repeatedly ranked highly as shown in
Figure~\ref{fig:probend_results}, and there are cases for which they were the
winners of the tournament.

@@

The distributions of the normalised rank in probabilistic ending tournaments,
Figure~\ref{fig:probend_results}, are less variant than those of noisy
tournaments. The medians are lower than 0.1 and the distributions are skewed
towards 0. Though the large difference between the means and the medians
indicates some outliers, the strategies have overall performed well in the
probabilistic ending tournaments that they participated.

In tournaments with both noise and an unspecified number of turns several of
the top ranked strategies are strategies that were highly ranked in noisy
tournaments. This demonstrates that, $e$, $\pi$,
$\phi$ and the Retaliate family are robust in noisy environments regardless the
number of turns. However, strategies from the top ranks in probabilistic
ending tournaments did not rank highly here. The distributions of the top scored
strategies are shown in Figure~\ref{fig:noisy_probend_results}. There is no
statistical difference between the distributions (this has been tested using
an ANOVA test). Thus, on average these strategies had the same performances.

@@

The top ranks include strategies that have previously mentioned. The top ranks
are overtaken by the set of Retaliate strategies followed by BackStabber and
DoubleCrosser. DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
stochastic  &  If a strategy is stochastic & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
memory usage &  The memory size of a strategy divided by the number of turns & memory size from~\cite{axelrodproject} & float & 0 &  1 \\
SSE & A measure of how far a strategy is from extortionate behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in the tournament  & result summary  & float & 0 & 1\\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in the tournament  & result summary  & float & 0 & 1\\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in the tournament  & result summary  & float & 0 & 1\\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in the tournament  & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median  & result summary  & float & 0 & 1\\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & result summary  & float & 0 & 1 \\
$C_r$ & The cooperating ratio of a strategy & result summary  & float & 0 & 1 \\
$CC$ to $C$ rate & The probability a strategy will cooperate after a mutual cooperation & result summary  & float & 0 & 1\\
$CD$ to $C$ rate & The probability a strategy will cooperate after being betrayed by the opponent & result summary  & float & 0 & 1 \\
$DC$ to $C$ rate & The probability a strategy will cooperate after betraying the opponent & result summary  & float & 0 & 1 \\
$DD$ to $C$ rate & The probability a strategy will cooperate after a mutual defection & result summary  & float & 0 & 1 \\
$p$ & The probability of a player's action being flip at each interaction & trial summary & float & 0 & 1 \\
$n$ & The number of turns & trial summary & integer & 1 & 200 \\
$e$ & The probability of a match ending in the next turn & trial summary & float & 0 & 1 \\
$N$ & The number of strategies in the tournament & trial summary & integer & 3 & 195 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}
    \label{table:manual_measures}
\end{table}

Axelrod-Python makes use of classifiers to classify strategies according to
various dimensions. The classifiers considered in this work are the stochastic
classifier, the make use of length and game classifiers. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage measure is calculated as the
memory size of strategy (which is specified in the strategies implementation
in~\cite{axelrodproject}) divide by the number of turns. For tournaments with a
probabilistic ending the number of turns has been not collected, so the memory
usage measure is not used for probabilistic ending tournaments.
The SSE is a measure of extortionate behaviour introduced in~\cite{Knight2019}.
A value of 1 indicates no extortionate behaviour at all whereas a value of 0
indicates that a strategy is always trying to extortion it's opponent. The rest
of the factors considered are the $CC$ to $C$, $CD$ to $C$, $DC$ to $C$, and
$DD$ to $C$ rates as well as cooperating ratio of a strategy. The minimum,
maximum, medium and median cooperating ratios of each tournament are also
included, and finally the number of turns, the number of strategies and the
probabilities of noise and the game ending are also included.

Initially, the effect of these measures has on a strategy's success is evaluated
based on the correlation coefficient. The correlation coefficient between them,
the normalised rank and the median score are given by
Table~\ref{table:correlations}. Note that the correlation for the classifiers is
not included because they are binary variables and they will be evaluated by a
different method. The correlation coefficients for all the measures in
Table~\ref{table:manual_measures} against each have also been calculated and apd
graphical representation can be found in the Appendix~\ref{app:correlations}.

@@

In standard tournaments the measures  $CC$ to $C$, $C_r$, $C_r / C_{\text{max}}$
and the cooperating ratio compared to $C_{\text{median}}$ and $C_{\text{mean}}$
have a moderate negative effect on the normalised rank, and a moderate positive
on the median score. The SSE error and the $DD$ to $C$ have the opposite
effects. Thus, in standard tournaments behaving cooperative corresponds to a
more successful performance. Even though being nice pays off,
that's not true against defective strategies. Cooperating after a mutual
defection lowers a strategy's success.
Figure~\ref{fig:rates_of_winners_in_standard_tournaments} confirms that the
winners of standard tournaments always cooperate after a mutual cooperation and
almost always defected after a mutual defection.

@@

In noisy probabilistic ending tournaments and in over all the tournaments' results,
the pdonly measures that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

A random forest approach is applied to the data to predict the cluster to which
a strategy's performance has been assigned to. The random forest method
constructs many individual decision trees and the predictions from all trees are
pooled to make the final prediction. The random forest models are trained on a
training set of 70\% of the tournaments results. The accuracy of each model
based on $R^2$ are given by Table~\ref{table:accuracy_random_forest}. The out of
the bag error~\cite{hastie2005} has also been calculated. The models fit well, and
a high value of both the accuracy measure on the test data and the
OOB error indicate that the model ispdf not over fitting.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.5\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    Tournament type & $R^2$ training data &  $R^2$ test data  & $R^2$ OBB score\\
    \midrule
    standard                   & 0.998545  & 0.989890 & 0.982331\\
    noisy                      & 0.996677  & 0.950572 & 0.935383\\
    probabilistic ending       & 0.999592  & 0.995128 & 0.992819 \\
    noisy probabilistic ending & 0.990490  & 0.813905 & 0.791418\\
    over \numberofalltournaments tournaments & 0.993396 & 0.913409 & 0.898059 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Accuracy metrics for random forest models.}
    \label{table:accuracy_random_forest}
\end{table}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_standard.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for standard tournaments.}\label{fig:mean_median_std}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_noisy.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for noisy tournaments.}\label{fig:mean_median_noisy}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_probend.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for probabilistic ending tournaments.}\label{fig:mean_median_probend}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_probend_noisy.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for noisy probabilistic ending tournaments.}\label{fig:mean_median_probend_noisy}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_overall.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for over all the tournaments.}\label{fig:mean_median_overall}
\end{figure}

In this section the effect of several measures, regarding a strategy's behaviour
and the tournament, on it's performance were presented. This was done using
two approaches. pCorrelation coefficients and a random forest
analysis. The results of these are discussed in the following section.

@@

This manuscript has explored the performance of 194 strategies of the Iterated
Prisoner's Dilemma inp a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament types, such as
standard, noisy and probabilistic ending tournaments, dominant strategies exist
there is not a single dominant strategy that is set to do if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators.

>>>>

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Today more than 200 strategies exist in the literature and several
tournaments, excluding Axelrod's, have been undertaken~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012}.

@@

Another set of IPD strategies that have received a lot of attention
are the zero-determinants~\cite{Press2012}. By
forcing a linear relationship between the payoffs the zero-determinant
strategies can ensure that they will never receive less than their opponents.
The
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire''.
Zero-determinant strategies are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including zero-determinants, was ran. None of the zero-determinant
strategies ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only a select pieces of work have been discussed, the IPD is
rich, and new strategies and competitions are being published every year.
The question, however, still remains the same: what is the best way to play the
game?

This manuscript uses an open source package called
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments, using as many strategies as possible from the literature. The aim
is to evaluate the performance of 194 strategies in IPD tournaments, and
futhermore, explore the factors of their success. This is done not only for
standard round robin tournaments, but also for tournaments with noise and
tournaments with a probabilistic ending. The different tournament types as well
as the data collection are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performing strategies
for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}.

This manuscripts uses several parameters. These are introduced in the following
sections, however, the full set of parameters and their definitions are given in
Appendix~\ref{app:parameters}. All the data is available at~\cite{data}.

@@

The result summary, Table~\ref{table:output_result}, has \(N\) number of rows
because each row contains information for each strategy that participated in the
tournament. The information includes the strategy's rank, median score, the rate
with which the strategy cooperated $(C_r)$, its match win count and the probability that
the strategy cooperated in the opening move. Moreover, the probabilities of a strategy
being in any of the four states ($CC, CD, DC, DD$), and the rate of which the
strategy cooperated after each state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h!]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result of a single tournament.}\label{table:output_result}
\end{table}

@@

Each strategy participated in multiple tournaments of the same type (on average 5180).
For example Tit For Tat has participated in a total of 5114
tournaments of each type. The strategy's normalised rank distribution in these
is given in Figure~\ref{fig:tit_for_tat_r_distribution}. A value of \(r =
0\) corresponds to a strategy winning the tournament where a value of
\(r = 1\) corresponds to the strategy coming last. Because of the strategies'
multiple entries their performance is evaluated based on the
\textbf{median normalised rank} denoted as \(\bar{r}\).

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy,
$e$, $\pi$ and $\phi$ (all from~\cite{axelrodproject}). The Retaliate and
Limited Retaliate strategies are implemented in~\cite{axelrodproject} by the
same contributor. The strategies are designed to defect if the opponent has
tricked the strategy more often than \(x\%\) of the times that the strategy has
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are bimodal.
This indicates that although the top ranked strategies mainly performed
well, there are several tournaments that they ranked in the bottom half.

@@

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever are ranked
$14^{\text{th}}$ and $15^{\text{th}}$. These strategies are actually the same;
they will defect until their opponent defect, then they will cooperate until the
end. Both strategies have repeatedly ranked highly as shown in
Figure~\ref{fig:probend_results}, and there are cases for which they were the
winners of the tournament.

@@

The distributions of the normalised rank in probabilistic ending tournaments,
shown in
Figure~\ref{fig:probend_results}, are less variant than those of noisy
tournaments. The medians are lower than 0.1 and the distributions are skewed
towards 0. Though the large difference between the means and the medians
indicates some outliers, the strategies have overall performed well in the
probabilistic ending tournaments that they participated.

In tournaments with both noise and an unspecified number of turns several of
the top ranked strategies are strategies that were highly ranked in noisy
tournaments. This demonstrates that, $e$, $\pi$,
$\phi$ and the Retaliate family are robust in noisy environments regardless of the
number of turns. However, strategies from the top ranks in probabilistic
ending tournaments did not rank highly here. The distributions of the top scored
strategies are shown in Figure~\ref{fig:noisy_probend_results}. There is no
statistical difference between the distributions (this has been tested using
an ANOVA test). Thus, on average these strategies had the same performances.

@@

The top ranks include strategies that have been previously mentioned. The top ranks
are overtaken by the set of Retaliate strategies followed by BackStabber and
DoubleCrosser. DoubleCrosser performed well in standard tournaments and the
strategy is just an extension of BackStabber. PSO Gambler and Evolved HMM 5 are
trained strategies introduced in~\cite{Harper2017} and Nice Meta Winner and NMWE
Memory One are strategies based on teams. Grudger is a strategy from Axelrod's
original tournament and Forgetful Fool Me Once is based on the same approach as
Grudger.

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
stochastic  &  If a strategy is stochastic & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
memory usage &  The memory size of a strategy divided by the number of turns & memory size from~\cite{axelrodproject} & float & 0 &  1 \\
SSE & A measure of how far a strategy is from ZD behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in the tournament  & result summary  & float & 0 & 1\\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in the tournament  & result summary  & float & 0 & 1\\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in the tournament  & result summary  & float & 0 & 1\\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in the tournament  & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & result summary  & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median  & result summary  & float & 0 & 1\\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & result summary  & float & 0 & 1 \\
$C_r$ & The cooperating ratio of a strategy & result summary  & float & 0 & 1 \\
$CC$ to $C$ rate & The probability a strategy will cooperate after a mutual cooperation & result summary  & float & 0 & 1\\
$CD$ to $C$ rate & The probability a strategy will cooperate after being betrayed by the opponent & result summary  & float & 0 & 1 \\
$DC$ to $C$ rate & The probability a strategy will cooperate after betraying the opponent & result summary  & float & 0 & 1 \\
$DD$ to $C$ rate & The probability a strategy will cooperate after a mutual defection & result summary  & float & 0 & 1 \\
$p$ & The probability of a player's action being flip at each interaction & trial summary & float & 0 & 1 \\
$n$ & The number of turns & trial summary & integer & 1 & 200 \\
$e$ & The probability of a match ending in the next turn & trial summary & float & 0 & 1 \\
$N$ & The number of strategies in the tournament & trial summary & integer & 3 & 195 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}
    \label{table:manual_measures}
\end{table}

Axelrod-Python makes use of classifiers to classify strategies according to
various dimensions. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage measure is calculated as the
memory size of strategy (which is specified in the strategies implementation
in~\cite{axelrodproject}) divide by the number of turns. For tournaments with a
probabilistic ending the number of turns has been not collected, so the memory
usage measure is not used for probabilistic ending tournaments.
The SSE is a measure of extortionate behaviour introduced in~\cite{Knight2019}.
A value of 1 indicates no extortionate behaviour at all whereas a value of 0
indicates that a strategy is always trying to extortion it's opponent. The rest
of the factors considered are the $CC$ to $C$, $CD$ to $C$, $DC$ to $C$, and
$DD$ to $C$ rates as well as cooperating ratio of a strategy. The minimum,
maximum, medium and median cooperating ratios of each tournament are also
included, and finally the number of turns, the number of strategies and the
probabilities of noise and the game ending are also included.

Table~\ref{table:correlations} shows the correlation coefficients between the
measures of Table~\ref{table:manual_measures} the median score and the median
normalised rank. Note that the correlation for the classifiers is
not included because they are binary variables and they will be evaluated using a
different method. The correlation coefficients for all the measures in
Table~\ref{table:manual_measures} against themselves have also been calculated and a
graphical representation can be found in the Appendix~\ref{app:correlations}.

@@

In standard tournaments the measures  $CC$ to $C$, $C_r$, $C_r / C_{\text{max}}$
and the cooperating ratio compared to $C_{\text{median}}$ and $C_{\text{mean}}$
have a moderate negative effect on the normalised rank, and a moderate positive
on the median score. The SSE error and the $DD$ to $C$ have the opposite
effects. Thus, in standard tournaments behaving cooperatively corresponds to a
more successful performance. Even though being nice pays off,
that's not true against defective strategies. Cooperating after a mutual
defection lowers a strategy's success.
Figure~\ref{fig:rates_of_winners_in_standard_tournaments} confirms that the
winners of standard tournaments always cooperate after a mutual cooperation and
almost always defects after a mutual defection.

@@

In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only measures that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

A random forest approach is applied to the data to predict the cluster to which
a strategy's performance has been assigned to. The random forest method
constructs many individual decision trees and the predictions from all trees are
pooled to make the final prediction. The random forest models are trained on a
training set of 70\% of the tournaments results. The accuracy of each model
based on $R^2$ are given by Table~\ref{table:accuracy_random_forest}. The out of
the bag error~\cite{hastie2005} has also been calculated. The models fit well, and
a high value of both the accuracy measure on the test data and the
OOB error indicate that the model is not over fitting.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.5\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    Tournament type & $R^2$ training data &  $R^2$ test data  & $R^2$ OOB score\\
    \midrule
    standard                   & 0.998545  & 0.989890 & 0.982331\\
    noisy                      & 0.996677  & 0.950572 & 0.935383\\
    probabilistic ending       & 0.999592  & 0.995128 & 0.992819 \\
    noisy probabilistic ending & 0.990490  & 0.813905 & 0.791418\\
    over \numberofalltournaments tournaments & 0.993396 & 0.913409 & 0.898059 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Accuracy metrics for random forest models.}
    \label{table:accuracy_random_forest}
\end{table}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_standard.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for winners of standard tournaments.}\label{fig:mean_median_std}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_noisy.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for winners of noisy tournaments.}\label{fig:mean_median_noisy}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_probend.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for winners of probabilistic ending tournaments.}\label{fig:mean_median_probend}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_probend_noisy.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for winners of noisy probabilistic ending tournaments.}\label{fig:mean_median_probend_noisy}
\end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.5\textwidth]{../images/compared_to_mean_median_overall.pdf}
    \caption{Distributions of \(C_r / C_{\text{median}}\)
    and \(C_r / C_{\text{median}}\) for winners of over all the tournaments.}\label{fig:mean_median_overall}
\end{figure}

In this section the effect of several measures, regarding a strategy's behaviour
and the tournament, on it's performance were presented. This was done using
two approaches. Correlation coefficients and a random forest
analysis. The results of these are discussed in the following section.

@@

This manuscript has explored the performance of 194 strategies of the Iterated
Prisoner's Dilemma in a large number of computer tournaments. The results of
the analysis demonstrated that, although for specific tournament type, such as
standard, noisy and probabilistic ending tournaments, dominant strategies exist
there is not a single dominant strategy if the environments
vary. Moreover, a strategy with a theory of mind should aim to adapt its behaviour
based on the mean and median cooperators.
